---
title: "Using the functionshelp Package"
author: "Benetti Gregorio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using_functionshelp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(functionshelp)
```

dt1 <- task1_filter_counts()
df1 <- task1_df_filter_counts()
print(head(dt1))
print(head(df1))

dt2 <- task2_add_qc_flags()
df2 <- task2_df_add_qc_columns()
print(head(dt2))
print(head(df2))

dt3 <- task3_join_and_index()
df3 <- task3_df_join_metadata()
print(head(dt3))
print(head(df3))

dt4 <- task4_annotate_counts()
df4 <- task4_df_annotate_counts()
print(head(dt4))
print(head(df4))

dt5 <- task5_classify_labs()
df5 <- task5_df_classify_labs()
print(head(dt5))
print(head(df5))

dt6 <- task6_match_labs_vitals()
df6 <- task6_df_nearest_vitals()
print(head(dt6))
print(head(df6))

dt7 <- task7_filter_genomic_peaks()
df7 <- task7_df_slice_peaks()
print(head(dt7))
print(head(df7))

dt8 <- task8_summarize_gene_stats()
df8 <- task8_df_summary_stats()
print(head(dt8))
print(head(df8))

dt9 <- task9_reshape_counts()
df9 <- task9_df_reshape_counts()
print(head(dt9))
print(head(df9))

dt10 <- task10_map_peaks_to_genes()
df10 <- task10_df_map_peaks()
print(head(dt10))
print(head(df10))

dt11 <- task11_map_variants_to_genes()
df11 <- task11_df_map_snps()
print(head(dt11))
print(head(df11))

dt12 <- task12_combine_cohorts()
df12 <- task12_df_combine_cohorts()
print(head(dt12))
print(head(df12))





